Comparison of dimensionality reduction schemes for parallel global optimization algorithms \thanks{The study was supported by the Russian Science Foundation, project No 16-11- 10150}

### 2018-04-13 2018/RSD/MultiEvolvents/RSD.tex
RSD: shortened english version

\begin{document}
\setlength{\abovedisplayskip}{3pt}
\setlength{\belowdisplayskip}{3pt}
%
\title{Comparison of dimensionality reduction schemes for parallel global optimization
algorithms
\thanks{The study was supported by the Russian Science Foundation, project No 16-11-
10150}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Konstantin Barkalov \and
Vladislav Sovrasov \and
Ilya Lebedev}
%
\authorrunning{K. Barkalov et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{
Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, Russia
\email{konstantin.barkalov@itmm.unn.ru}\\
\email{sovrasov.vlad@gmail.com}\\
\email{ilya.lebedev@itmm.unn.ru}\\
\url{http://hpc-education.unn.ru/}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
This work considers a parallel algorithms for solving multi-extremal optimization problems.
Algorithms are developed within the framework of the information-statistical approach and
implemented in a parallel solver Globalizer. The optimization problem is solved by reducing
the multidimensional problem to a set of joint one-dimensional problems that are solved in
parallel. Five types of Peano-type space-filling curves are employed to reduce dimension. The
results of computational experiments carried out on several hundred test problems are discussed.

@@

%\end{Russian}
%------------------------------------------------------------------------------
\section{Statement of Multidimensional Global Optimization Problem}
In this paper, the core class of optimization problems, which can be solved using
Globalizer, is formulated. This class involves the multidimensional global
optimization problems without constraints, which can be defined in the following way:
\begin{equation}
\label{eq:task}
\begin{array}{cr}\\
  \varphi(y^*)=\min\{\varphi(y):y\in D\}, \\
  D=\{y\in \mathbb{R}^N:a_i\leq y_i\leq{b_i}, 1\leq{i}\leq{N}\}
\end{array}
\end{equation}
with the given boundary vectors  $a$ and  $b$. It is supposed, that the objective function
\(\varphi(y)\) satisfies the Lipschitz condition
\begin{equation}
\label{eq:lip}
|\varphi(y_1)-\varphi(y_2)|\leq L\Vert y_1-y_2\Vert,y_1,y_2\in D,
\end{equation}
where \(L>0\) is the Lipschitz constant, and \(||\cdot||\) denotes the norm in \(\mathbb{R}^N\)
space.
\par
Usually, the minimized function \(\varphi(y)\) is defined as a computational procedure,
according to which the value \(\varphi(y)\) can be calculated for any vector \(y\in D\)
(let us further call such a calculation \textit{a trial}). It is supposed that this procedure
is a time-consuming one. As a result, the overall time of solving the optimization
problem (\ref{eq:task}) is determined, first of all by the number of executed trials.
It should also be noted that the requirement of the Lipschitz condition (\ref{eq:lip})
is highly important, since an estimate of the global minimum can be constructed on the
basis of a finite number of computed values of the optimized function only in this case .

@@

The reduction of the multidimensional problems to the one-dimensional ones using the Peano
curves has such important properties as the continuity and the uniform boundedness of the
function differences for limited variation of argument. However, a partial loss of information on
the nearness of the points in the multidimensional space takes place since a point $x \in [0,1]$
has the left and the right neighbors only while the corresponding point $y(x) \in D \subset R^N$
has the neighbors in $2N$ directions. As a result, when using the mappings like Peano curve the
images $y',\ y'',$ which are close to each other in the $N$-dimensional space can correspond to
the preimages $x',\ x'',$ which can be far away from each other in the interval $[0,1]$. This
property results in the excess computations since several limit points $x',\ x''$ of the trial
sequence generated by the index method in the interval $[0,1]$ can correspond to a single limit
point $y$ in the $N$-dimensional space.

One of the possible ways to overcome this disadvantage consists in using the multiple mappings
\begin{equation}%\label{eq:142}
Y_L(x)=\left\{y^0(x),\ y^1(x),...,\ y^L(x)\right\}
\end{equation}
instead of single Peano curve $y(x)$ (see \cite{Strongin1991,Strongin1992,strSergGO}).

@@

In Fig.~\ref{fig:shifted_ev} the image of the interval $[0,1]$ obtained by the curve $y^0(x),\
x\in [0,1],$ is shown as the dashed line. Since the hypercube $D$ from (\ref{eq:task}) is
included in the common part of the family of hypercubes $D_l$ (the boundaries of hypercube
$D$ are highlighted in Fig.~\ref{fig:shifted_cube}), having introduced an additional constraint
function
\begin{equation}\label{6_g0}
g_0(y)=\max\left\{\left|y_i\right| - 2^{-1}:\ 1\leq i\leq N\right\},
\end{equation}
one can present the initial hypercube $D$ in the form
\[
D=\left\{y^l(x):\; x\in [0,1],\ g_0(y^l(x))\leq 0 \right\},\ 0\leq l \leq L,
\]
i.e., $g_0(y) \leq 0$ if $y\in D$ and $g_0(y)>0$ otherwise. Consequently, any point $y \in D$
has its own preimage $x^l \in [0,1]$ for each mapping $y^l(x),\ 0\leq l\leq L$.

@@

\begin{figure}[ht]
    \centering
    \subfloat[Two
mappings]{{\includegraphics[width=.6\textwidth]{pictures/shifted.pdf}}\label{fig:shifted_ev}
}
    \subfloat[Hypercubes
$D_l$]{{\includegraphics[width=.4\textwidth]{pictures/shifted_cube.png}}\label{fig:shifted_c
ube}}
    \caption{Multiple mappings}
\end{figure}

%------------------------------------------------------------------------------
\subsection{Rotated evolvents}
The application of the scheme for building the multiple evolvents (hereinafter called the shifted
evolvents or $S$-evolvents) described in Subsection \ref{sec:shifted} allows to preserve the
information on the nearness of the points in the multidimensional space and, therefore, to
provide more precise (as compared to a single evolvent) estimate of Lipschitz constant in the
search process. However, this approach has serious restrictions, which narrow the applicability
of the parallel algorithms, designed on the base of the $S$-evolvents (see te end of the section
\ref{sec:seq_comp}).

To overcome the $S$-evolvent and to preserve the information on the nearness of the points in
the $N$-dimensional space, a novel scheme of building of the multiple mappings is proposed.
The building of a set of Peano curves not by the shift along the main diagonal of the hypercube
but by rotation of the evolvents around the coordinate origin is a distinctive feature of the
proposed scheme \cite{Gergel2009}. In the initial non-rotated mapping for close points $y', y''$
in the multidimensional space their preimages  $x', x''$ in the interval $[0,1]$ can be far away
from each other. In the rotated scheme there exists a mapping $y^i(x)$ according to which
preimages $x', x''$ will be located nearer. The evolvents generated according to this scheme are
called the rotated evolvents or $R$-evolvents.

In Fig.~\ref{6_fig_9} two evolvents being the approximations to Peano curves for the case
$N=2$ are presented as an illustration.
Taking into account the initial mapping, one can conclude that current implementation of the
method allows to build up to $N(N-1)+1$ evolvents for mapping the $N$-dimensional domain
onto the corresponding one-dimensional intervals. Moreover, the additional constraint  $g_0(y)
\leq 0$ with $g_0(y)$ from (\ref{6_g0}), which arises in shifted evolvents, is absent. This
method for building a set of mappings can be ``scaled'' easily to obtain more evolvents (up to
$2^N$) if necessary.

\begin{figure}[t]
  \centering
  \includegraphics[width=0.6\linewidth]{pictures/rotated.pdf}
  \caption{Two rotated evolvents on the same plane}
  \label{6_fig_9}
\end{figure}

@@

The methods of constructing the evolvents considered in the previous paragraphs produce the
curve $y(x)$, which in not a smooth one (see Fig.~\ref{fig:shifted_ev}). The absence of
smoothness may affect the properties of the reduced one-dimensional function $\varphi(y(x))$
adversely since a smooth curve reflect the information on the growth/decay of the initial
function better. On the basis of initial algorithm of constructing the non-smooth evolvent, a
generalized algorithm allowing constructing a smooth space-filling curve has been
proposed~\cite{Goryachih2017}. As an illustration, a smooth evolvent for the two-dimensional
case is presented in Fig.~\ref{fig:smooth}.
An increased computational complexity (several times as compared to the piecewise-linear
curves) is a disadvantage of the smooth evolvent. This caused by computing of the nonlinear smooth
functions. At that, the number of smooth intervals and the difficulty of the
computations of the curve increases with increasing accuracy of approximation and
space dimension.

@@

\begin{figure}[ht]
    \centering
    \subfloat[Smooth
evolvent]{{\includegraphics[width=.6\textwidth]{pictures/smooth.pdf}}\label{fig:smooth}}
    \subfloat[Numerical non-univalent
evolvent]{{\includegraphics[width=.4\textwidth]{pictures/noninjective.pdf}}\label{fig:noninje
ctive}}
    \caption{Different evolvents}
\end{figure}

@@

In order to evaluate the efficiency of an algorithm on a given set of 100 problems, we will use
the
operating characteristics \cite{grishaginClass}, which are defined by a set of
points on the \((K, P)\) plane where \(K\) is the average number of search trials
conducted before satisfying the termination condition when minimizing a function
from a given class, and \(P\) is the proportion of problems solved successfully.
If at a given \(K\), the operating characteristic of a method goes higher than one
from another method, it means that at fixed search costs, the former method has a
greater probability of finding the solution. If some value of \(P\) is fixed, and the
characteristic of a method goes to the left from that of another method, the former
method requires fewer resources to achieve the same reliability.

@@

\subfloat[$r=5.0$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS2d_same_r_opt_pt_op.pdf}}\label{fig:gkls2d_acc}}
    \subfloat[Minimal
$r$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS2d_opt_pt_op.pdf}}\label{fig:gkls2
d_opt}}
    \caption{Operating characteristics on GKLS 2d Simple class}
\end{figure}

@@

\subfloat[$r=4.5$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS3d_same_r_opt_pt_op.pdf}}\label{fig:gkls3d_acc}}
    \subfloat[Minimal
$r$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS3d_opt_pt_op.pdf}}\label{fig:gkls3
d_opt}}
    \caption{Operating characteristics on GKLS 3d Simple class}
\end{figure}

>>>>

\begin{document}
\setlength{\abovedisplayskip}{3pt}
\setlength{\belowdisplayskip}{3pt}
%
\title{Comparison of dimensionality reduction schemes for parallel global optimization
algorithms
\thanks{The study was supported by the Russian Science Foundation, project No 16-11-
10150}}
%
\titlerunning{Dimension reduction schemes in parallel GO}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Konstantin Barkalov \and
Vladislav Sovrasov \and
Ilya Lebedev}
%
\authorrunning{K. Barkalov et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{
Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, Russia
\email{konstantin.barkalov@itmm.unn.ru}\\
\email{sovrasov.vlad@gmail.com}\\
\email{ilya.lebedev@itmm.unn.ru}\\
\url{http://hpc-education.unn.ru/}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
This work considers a parallel algorithms for solving multi-extremal optimization problems.
Algorithms are developed within the framework of the information-statistical approach and
implemented in a parallel solver Globalizer. The optimization problem is solved by reducing
the multidimensional problem to a set of joint one-dimensional problems that are solved in
parallel. Five types of Peano-type space-filling curves are employed to reduce dimension. The
results of computational experiments carried out on several hundred test problems are discussed.

@@

%\end{Russian}
%------------------------------------------------------------------------------
\section{Statement of Multidimensional Global Optimization Problem}
In this paper, the core class of optimization problems, which can be solved using
Globalizer, is formulated. This class involves the multidimensional global
optimization problems without constraints, which can be defined in the following way:
\begin{equation}
\label{eq:task}
\begin{array}{cr}\\
  \varphi(y^*)=\min\{\varphi(y):y\in D\}, \\
  D=\{y\in \mathbb{R}^N:a_i\leq y_i\leq{b_i}, 1\leq{i}\leq{N}\}
\end{array}
\end{equation}
with the given boundary vectors  $a$ and  $b$. It is supposed, that the objective function
\(\varphi(y)\) satisfies the Lipschitz condition
\begin{equation}
\label{eq:lip}
|\varphi(y_1)-\varphi(y_2)|\leq L\Vert y_1-y_2\Vert,y_1,y_2\in D,
\end{equation}
where \(L>0\) is the Lipschitz constant, and \(||\cdot||\) denotes the norm in \(\mathbb{R}^N\)
space.
\par
Usually, the minimized function \(\varphi(y)\) is defined as a computational procedure,
according to which the value \(\varphi(y)\) can be calculated for any vector \(y\in D\)
(let us further call such a calculation \textit{a trial}). It is supposed that this procedure
is time-consuming.
%As a result, the overall time of solving the optimization
%problem (\ref{eq:task}) is determined, first of all by the number of executed trials.
%It should also be noted that the requirement of the Lipschitz condition (\ref{eq:lip})
%is highly important, since an estimate of the global minimum can be constructed on the
%basis of a finite number of computed values of the optimized function only in this case .

@@

One of the possible ways to overcome the negative effects of using a numerical
approximation of evolvent (it destroys the information about the neighbour points in $\mathbb{R}^N$ space,
see \cite{Strongin1992}) consists in using the multiple mappings
\begin{equation}%\label{eq:142}
Y_L(x)=\left\{y^0(x),\ y^1(x),...,\ y^L(x)\right\}
\end{equation}
instead of single Peano curve $y(x)$ (see \cite{Strongin1991,Strongin1992,strSergGO}).

@@

In Fig.~\ref{fig:shifted_ev} the image of the interval $[0,1]$ obtained by the curve $y^0(x),\
x\in [0,1],$ is shown as the dashed line. Since the hypercube $D$ from (\ref{eq:task}) is
included in the common part of the family of hypercubes $D_l$, having introduced an additional constraint
function
\begin{equation}\label{6_g0}
g_0(y)=\max\left\{\left|y_i\right| - 2^{-1}:\ 1\leq i\leq N\right\},
\end{equation}
one can present the initial hypercube $D$ in the form
\[
D=\left\{y^l(x):\; x\in [0,1],\ g_0(y^l(x))\leq 0 \right\},\ 0\leq l \leq L,
\]
i.e., $g_0(y) \leq 0$ if $y\in D$ and $g_0(y)>0$ otherwise. Consequently, any point $y \in D$
has its own preimage $x^l \in [0,1]$ for each mapping $y^l(x),\ 0\leq l\leq L$.

@@

\begin{figure}[ht]
    \centering
    \subfloat[Two shifted evolvents on the hypercubes $D_0$ and $D_1$]{{\includegraphics[width=.5\textwidth]{pictures/shifted.pdf}}\label{fig:shifted_ev}}
    %\subfloat[Hypercubes $D_l$]{{\includegraphics[width=.4\textwidth]{pictures/shifted_cube.png}}\label{fig:shifted_cube}}
    \subfloat[Two rotated evolvents on the same plane]{{\includegraphics[width=.5\textwidth]{pictures/rotated.pdf}}\label{6_fig_9}}
    \caption{Multiple mappings}
\end{figure}

%------------------------------------------------------------------------------
\subsection{Rotated evolvents}
The application of the scheme for building the multiple evolvents (hereinafter called the shifted
evolvents or $S$-evolvents) described in Subsection \ref{sec:shifted} allows to preserve the
information on the nearness of the points in the multidimensional space and, therefore, to
provide more precise (as compared to a single evolvent) estimate of Lipschitz constant in the
search process. However, this approach has serious restrictions, which narrow the applicability
of the parallel algorithms, designed on the base of the $S$-evolvents (see the end of the section
\ref{sec:seq_comp}).

To overcome complexity of the $S$-evolvent and to preserve the information on the nearness of the points in
the $N$-dimensional space, one more scheme of building of the multiple mappings was proposed.
The building of a set of Peano curves not by the shift along the main diagonal of the hypercube
but by rotation of the evolvents around the coordinate origin is a distinctive feature of the
proposed scheme \cite{Gergel2009}.
%In Fig.~\ref{6_fig_9} two evolvents being the approximations to Peano curves for the case
%$N=2$ are presented as an illustration.
Taking into account the initial mapping, one can conclude that current implementation of the
method allows to build up to $N(N-1)+1$ evolvents for mapping the $N$-dimensional domain
onto the corresponding one-dimensional intervals. Moreover, the additional constraint  $g_0(y)
\leq 0$ with $g_0(y)$ from (\ref{6_g0}), which arises in shifted evolvents, is absent. This
method for building a set of mappings can be ``scaled'' easily to obtain more evolvents (up to
$2^N$) if necessary.

%\begin{figure}[t]
%  \centering
%  \includegraphics[width=0.6\linewidth]{pictures/rotated.pdf}
%  \caption{Two rotated evolvents on the same plane}
%  \label{6_fig_9}
%\end{figure}

@@

The methods of constructing the evolvents considered in the previous paragraphs produce the
curve $y(x)$, which in not a smooth one (see Fig.~\ref{fig:shifted_ev}). The absence of
smoothness may affect the properties of the reduced one-dimensional function $\varphi(y(x))$
adversely since a smooth curve reflect the information on the growth/decay of the initial
function better. On the basis of initial algorithm of constructing the non-smooth evolvent, a
generalized algorithm allowing constructing a smooth space-filling curve has been
proposed~\cite{Goryachih2017}. As an illustration, a smooth evolvent for the two-dimensional
case is presented in Fig.~\ref{fig:smooth}.
An increased computational complexity (several times as compared to the piecewise-linear
curves) is a disadvantage of the smooth evolvent. This caused by computing of the nonlinear smooth
functions.
%At that, the number of smooth intervals and the difficulty of the
%computations of the curve increases with increasing accuracy of approximation and
%space dimension.

@@

\begin{figure}[ht]
    \centering
    \subfloat[Smooth
evolvent]{{\includegraphics[width=.6\textwidth]{pictures/smooth.pdf}}\label{fig:smooth}}
    \subfloat[Numerical non-univalent
evolvent]{{\includegraphics[width=.4\textwidth]{pictures/noninjective.pdf}}\label{fig:noninjective}}
    \caption{Different evolvents}
\end{figure}

@@

In order to evaluate the efficiency of an algorithm on a given set of 100 problems, we will use
the operating characteristics \cite{grishaginClass}, which are defined as a
curve, showing the dependency of number of solved problems vs the number of iterations.

@@

\subfloat[$r=5.0$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS2d_same_r_opt_pt_op.pdf}}\label{fig:gkls2d_acc}}
    \subfloat[Minimal
$r$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS2d_opt_pt_op.pdf}}\label{fig:gkls2d_opt}}
    \caption{Operating characteristics on GKLS 2d Simple class}
\end{figure}

@@

\subfloat[$r=4.5$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS3d_same_r_opt_pt_op.pdf}}\label{fig:gkls3d_acc}}
    \subfloat[Minimal
$r$]{{\includegraphics[width=.5\textwidth]{pictures/gklsS3d_opt_pt_op.pdf}}\label{fig:gkls3d_opt}}
    \caption{Operating characteristics on GKLS 3d Simple class}
\end{figure}
